Type-Safe Modular Parsing

### 2016-12-28 paper/sections/Overview.tex
Partly revised S3

\section{Overview}\label{sec:overview}

This section gives an overview of our library \name, and the problem that motivates our work. Basically, \name consists of four parts: underlying parsing technique, delegation mechanism encoded by open recursion, Object Algebras, and glue code of new combinators and utility functions. We start from Section \ref{subsec:overview-parsing}, which discusses the choice of parsing technique and how it affects modularity of parsers. Section \ref{subsec:overview-problem} demonstrates the goal of extending parsers together with ASTs in a semantic modular way, with both separate compilation and type-safe code reuse. Then we will see traditional parser combinators fail to achieve it because of hard-coded recursive calls. In Section \ref{subsec:overview-delegation}, we show how delegation can solve this problem and allow us to build extensible parsers. Finally, Section \ref{subsec:overview-oa} gives examples of using Object Algebras for more extensibility, including extension of operations and parsing multiple sorts of syntax.\haoyuan{TODO}

@@

\paragraph{\name} Our library is constructed by only around twenty lines of code. We have wrapped up the type of parsers
to make them modular and extensible, and we use our custom combinator \lstinline{<|>} for alternative composition of those
modular ``open parsers'' with longest match, built on \lstinline{|||}. Table~\ref{tab:parser} shows the API that \name provides for modular parsing.

@@

In the following text we are going to illustrate the problem that motivates our work.
Typical parsers defined by parser combinators follow the grammar
structure, they parse an input to build a corresponding AST. If the grammar is extended
with a new alternative, the corresponding parser has to be changed as
well. Thus the traditional style of developing parsers is not
extensible, and it is at odds with object-oriented programming ASTs,
which support the easy addition of new alternative classes. We
illustrate this with a simple expression language and extensions.

Suppose we want to parse a simple language of literals and
additions. The concrete syntax is as follows:

@@

This time we allow both parts of an addition to be expressions.
It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases:
%\huang{This is the first time to demonstrate code in this paper. Readers may not know what those combinators/types are. Refer to the library docs here? move the 'scala parsing library' section to the beginning as 'background'?}

@@

\noindent Note that we use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} in the rest of paper. As the combination of \inlinecode{pVar}, \inlinecode{pApp}, \inlinecode{pExpr} can parse expressions like \lstinline{"1+2"} correctly. At this point, one would like such code to be extensible in two dimensions, for example:
\begin{itemize}
\item \textbf{Challenge 1}: adding a new case (or rather, a new construct) to \lstinline{Expr};
\item \textbf{Challenge 2}: adding a new operation to \lstinline{Expr}.
\end{itemize}

\paragraph{Attempt to Extend the Parser} Now we firstly consider challenge 1, namely we are extending the syntax as well as the parser. Hence we introduce variables as a new case.

@@

The corresponding AST is extended as below.

@@

Since we already have the parser for literals and additions, we would like to build the new parser by reusing the old one. So we compose \inlinecode{pVar}, which parses variables, with the old parser. \inlinecode{pExtExpr} is the new function to parse expressions.

@@

Unfortunately, this extended parser does not work properly to parse all possible expressions defined by the new grammar. For example, it fails to parse \inlinecode{"1 + x"} which is obviously valid.

\paragraph{Problem: Hard-coded Recursive Calls} The critical problem of extending parsers directly is that the recursive calls in old parsers are hard-coded. Specifically, \inlinecode{pAdd} makes two recursive calls to parse two sub-expressions around \lstinline{"+"}. Before the extension, we implemented it by using \inlinecode{pExpr} for those recursive calls. It worked fine at that time because \inlinecode{pExpr} covers all cases of the old grammar. However, after the extension of variables, the proper function for such recursive calls should be \inlinecode{pExtExpr} instead of \inlinecode{pExpr}, because the new function \inlinecode{pExtExpr} includes the new extension.

@@

>>>>

\section{An Overview of \name}\label{sec:overview}

This section gives an overview of our library \name, motivates
additional problems that occur when parsing modular ASTs.  The first
problem is that traditional parsers using parser combinators use
hard-coded recursive calls. This problem is solved using
delegation-based techniques, which enable parsing OO ASTs.  The second
problem is how to achieve dual extensibility of ASTs (that is: how to
allow for both new language constructs and new operations over the
ASTs). This problem is solved using Object Algebras.

\begin{comment}
Basically, \name consists of four parts: underlying parsing technique, delegation mechanism encoded by open recursion, Object Algebras, and glue code of new combinators and utility functions. We start from Section \ref{subsec:overview-parsing}, which discusses the choice of parsing technique and how it affects modularity of parsers. Section \ref{subsec:overview-problem} demonstrates the goal of extending parsers together with ASTs in a semantic modular way, with both separate compilation and type-safe code reuse. Then we will see traditional parser combinators fail to achieve it because of hard-coded recursive calls. In Section \ref{subsec:overview-delegation}, we show how delegation can solve this problem and allow us to build extensible parsers. Finally, Section \ref{subsec:overview-oa} gives examples of using Object Algebras for more extensibility, including extension of operations and parsing multiple sorts of syntax.\haoyuan{TODO}
\end{comment}

@@

@@

In the following text we are going to illustrate the problems that
motivate our work via a minimalistic example inspired by Wadler's
Expression Problem~\cite{}.

\paragraph{Parsing a simple Arithmetic Language}
Suppose we want to parse a simple language with literals and
additions. The concrete syntax is as follows:
\bruno{Here's a good opportunity to both improve the paper and make it
more compact! This is the example that should be illustrated and
discussed at the end of Section 2. Then we can just create a figure to
refer to the code in Section 2 and refer again to the figure here. No
need for two different examples.}

@@

\noindent It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases:
%\huang{This is the first time to demonstrate code in this
%paper. Readers may not know what those combinators/types are. Refer
%to the library docs here? move the 'scala parsing library' section to
%the beginning as 'background'?}

@@

\noindent Note that we use \lstinline{Parser} as a type synonym for
\lstinline{PackratParser} in the rest of paper. 
The parser \inlinecode{pExpr} is able to parse 
expressions like \lstinline{"1+2"} correctly.

\paragraph{Dimensions of Extensibility}
At this point, one would like such code to be extensible in two dimensions, for example:
\begin{itemize}
\item \textbf{Extension 1}: adding a new case (or rather, a new construct) to \lstinline{Expr};
\item \textbf{Extension 2}: adding a new operation to \lstinline{Expr}.
\end{itemize}

\paragraph{Attempt to Extend the Parser} Now we first consider 
extension 1, namely we are extending the syntax as well as the parser. Hence we introduce variables as a new case.

@@

As in an OO solution to the Expression Problem, it is easy to extend
the corresponding AST:

@@

Unfortunatelly, extending the parser for the AST is not as easy.
Since we already have the parser for literals and additions, we would
like to build the new parser by reusing the old one. So we compose
\inlinecode{pVar}, which parses variables, with the old
parser. \inlinecode{pExtExpr} is the new function to parse
expressions.

@@

\noindent This extended parser does not work properly to parse all possible expressions defined by the new grammar. For example, it fails to parse \inlinecode{"1 + x"} which is obviously valid.

\paragraph{Problem: Hard-coded Recursive Calls} 
Typical parsers defined by parser combinators follow the grammar
structure, they parse an input to build a corresponding AST. If the grammar is extended
with a new alternative, the corresponding parser has to be changed as
well. Thus the traditional style of developing parsers is not
extensible, and it is at odds with object-oriented programming ASTs,
which support the easy addition of new alternative classes.
The critical problem
of extending parsers directly is that the recursive calls in old
parsers are hard-coded. Specifically, \inlinecode{pAdd} makes two
recursive calls to parse two sub-expressions around
\lstinline{"+"}. Before the extension, we implemented it by using
\inlinecode{pExpr} for those recursive calls. It worked fine at that
time because \inlinecode{pExpr} covers all cases of the old
grammar. However, after the extension of variables, the proper
function for such recursive calls should be \inlinecode{pExtExpr}
instead of \inlinecode{pExpr}, because the new function
\inlinecode{pExtExpr} includes the new extension.

@@

\bruno{Put the following text somewhere?}
\paragraph{\name} Our library is constructed by only around twenty lines of code. We have wrapped up the type of parsers
to make them modular and extensible, and we use our custom combinator \lstinline{<|>} for alternative composition of those
modular ``open parsers'' with longest match, built on \lstinline{|||}. Table~\ref{tab:parser} shows the API that \name provides for modular parsing.
